---
title: "hBEC RNA-Seq Analysis"
author: "MJR"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: journal
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = "../output/reports",
      knit_root_dir = ".."
    )
  })
---

# Setup & Configuration

## Load Libraries & Functions

```{r setup, include=FALSE}
# Set root to project directory (one level up from analysis/)
knitr::opts_knit$set(root.dir = here::here())
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

# Add this diagnostic chunk right after setup:
knitr::opts_chunk$set(
  echo = TRUE, 
  message = FALSE, 
  warning = FALSE,
  fig.width = 10,
  fig.height = 8
)
knitr::opts_knit$set(root.dir = here::here())

# Set seed for reproducibility
set.seed(4454)
```

```{r load-config}
# Source all configuration and function definitions
source("R/_load_all.R")

# Additional libraries specific to RNA-seq analysis
library(DESeq2)
library(edgeR)
library(statmod)
library(variancePartition)
library(BiocParallel)
library(EnhancedVolcano)
library(ComplexHeatmap)
library(circlize)
library(eulerr)
library(ggrepel)
library(rvest)
library(factoextra)

cat("✓ RNA-seq analysis environment loaded\n")
```

---

# Data Loading & Preparation

## Load Raw Data Files

```{r load-raw-data}
# Load raw counts
raw_seq <- read_csv(here::here(PATH_DATA_RAW, "30_999361309_genes_v2.csv"))

# Load metadata
metadata_MJR1_191 <- read_csv(here::here(PATH_DATA_RAW, "2023_MJR1_191_metadata.csv"))

# Subset MJR samples only
mjr_seq <- raw_seq %>%
  select(Geneid, starts_with("MJR")) %>%
  rename_all(~gsub("MJR-", "MJR_", .))

cat("✓ Raw data loaded\n")
cat("  - Genes:", nrow(mjr_seq), "\n")
cat("  - Samples:", ncol(mjr_seq) - 1, "\n")
```

## Filter Low-Abundance Genes

```{r filter-genes}
# Remove samples with 90% or more zeroes
mjr_seq <- mjr_seq[, colMeans(mjr_seq != 0) > 0.1]

# Apply background filtering using function
mjr_seq_filtered <- filter_genes(mjr_seq, background_threshold = GENE_BACKGROUND_THRESHOLD)

# Remove samples for incomplete donor (DD007T)
drop_cols <- c("MJR_146", "MJR_148", "MJR_152", "MJR_154", "MJR_158", 
               "MJR_160", "MJR_162", "MJR_164", "MJR_166")
mjr_seq_filtered <- mjr_seq_filtered %>% select(-any_of(drop_cols))

cat("✓ Gene filtering complete\n")
cat("  - Retained:", nrow(mjr_seq_filtered), "genes\n")
cat("  - Removed:", nrow(mjr_seq) - nrow(mjr_seq_filtered), "low-abundance genes\n")
cat("  - Final sample count:", ncol(mjr_seq_filtered) - 1, "\n")
```

## Prepare Metadata

```{r prepare-metadata}
# Subset metadata to match filtered samples
mjr_seq_samples <- colnames(mjr_seq_filtered)[-1]  # Exclude Geneid column

meta_data_subset <- metadata_MJR1_191 %>%
  filter(SAMPLEID %in% mjr_seq_samples) %>%
  select(SAMPLEID, EXPOSURE, SEX, PATIENTCODE) %>%
  mutate(
    EXPOSURE = gsub(" ", "_", EXPOSURE),
    EXPOSURE = gsub("Red_Oak_(5|25)", "RedOak_\\1", EXPOSURE),
    EXPOSURE = factor(EXPOSURE, levels = unique(EXPOSURE)),
    SEX = factor(SEX),
    PATIENTCODE = factor(PATIENTCODE)
  ) %>%
  arrange(match(SAMPLEID, mjr_seq_samples))  # Match order to seq data

# Set reference level
meta_data_subset$EXPOSURE <- relevel(meta_data_subset$EXPOSURE, ref = "PBS_Control")

cat("✓ Metadata prepared\n")
cat("  - Samples:", nrow(meta_data_subset), "\n")
cat("  - Exposures:", length(levels(meta_data_subset$EXPOSURE)), "\n")
cat("  - Donors:", length(unique(meta_data_subset$PATIENTCODE)), "\n")
```

---

# Quality Control & Exploratory Analysis

## PCA Plot

```{r pca-analysis, fig.width=10, fig.height=8}
# Prepare data for PCA
seq_transposed <- mjr_seq_filtered %>%
  pivot_longer(cols = -Geneid, names_to = "Sample", values_to = "Expression") %>%
  pivot_wider(names_from = Geneid, values_from = Expression) %>%
  left_join(meta_data_subset, by = c("Sample" = "SAMPLEID")) %>%
  column_to_rownames(var = "Sample")

# Convert to matrix for normalization
counts_matrix <- as.matrix(t(seq_transposed[, 1:(ncol(seq_transposed) - 3)]))

# Create DGEList and normalize
dataset <- DGEList(counts_matrix, group = seq_transposed$PATIENTCODE)
dataset <- calcNormFactors(dataset)
lcpm <- cpm(dataset, log = TRUE)

# Run PCA
pca_res <- prcomp(t(lcpm), center = TRUE, scale = TRUE)

# Plot with factoextra
fviz_pca_ind(
  pca_res,
  label = "none",
  habillage = seq_transposed$EXPOSURE,
  palette = color_exposure,
  addEllipses = TRUE,
  ellipse.type = 'convex'
) +
  ggtitle("PCA - Sample Clustering by Exposure") +
  theme(
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12),
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
    panel.border = element_rect(fill = NA, color = "black", size = 0.3),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_blank()
  )

save_plot("PCA_exposure_clustering.png", last_plot(), width = 10, height = 8)
```

## Sample Correlation Dendrogram

```{r dendrogram, eval=TRUE}
# Dendrogram analysis for sample QC
# Set eval=TRUE to run this during full QC

lcpm_combined <- as.data.frame(cbind(seq_transposed[, 1:2], t(lcpm)))
exposure_list <- as.list(unique(as.character(lcpm_combined$EXPOSURE)))

# Create dendrograms for each exposure group
dendrograms <- lapply(exposure_list, function(exposure) {
  tmp <- lcpm_combined %>% filter(EXPOSURE == exposure)
  names <- rownames(tmp)
  
  tmp <- tmp %>%
    select(-c(1:2)) %>%
    sapply(as.numeric) %>%
    as.data.frame(row.names = names)
  
  tmp_dend <- cor(t(tmp))
  tmp_dend <- as.dendrogram(hclust(as.dist(1 - tmp_dend)))
  
  par(mar = c(4, 1, 1, 12))
  plot(tmp_dend, main = as.character(exposure), horiz = TRUE) +
    abline(v = 0.25, lty = 2)
})

cat("✓ Dendrogram analysis complete\n")
```

---

# Normalization & Transformation

## Normalize Counts

```{r normalize-counts, cache=TRUE}
# Median-of-ratios normalization
counts_matrix <- as.matrix(mjr_seq_filtered[, -1]) + 1  # Add pseudocount

normalized_counts <- normalize_counts(counts_matrix)
normalized_counts <- as.data.frame(normalized_counts)
normalized_counts$Geneid <- mjr_seq_filtered$Geneid

# Reorder to put Geneid first
normalized_counts <- normalized_counts[, c("Geneid", setdiff(names(normalized_counts), "Geneid"))]

cat("✓ Median-of-ratios normalization complete\n")

# Export
write_csv(normalized_counts, here::here(PATH_DATA_PROCESSED, "MJR_seq_normalized.csv"))
```

## Variance Stabilizing Transformation

```{r vst-transform, cache=TRUE}
# Create DESeq2 object for VST
dds <- DESeqDataSetFromMatrix(
  countData = counts_matrix,
  colData = data.frame(condition = rep(1, ncol(counts_matrix))),
  design = ~ 1
)

# Estimate size factors and apply VST
dds <- estimateSizeFactors(dds)
vst_counts <- vst(dds, blind = TRUE)
vst_matrix <- assay(vst_counts)

# Convert to data frame
vst_counts_df <- as.data.frame(vst_matrix)
vst_counts_df$Geneid <- mjr_seq_filtered$Geneid
vst_counts_df <- vst_counts_df[, c("Geneid", setdiff(names(vst_counts_df), "Geneid"))]

cat("✓ VST transformation complete\n")

# Export
write_csv(vst_counts_df, here::here(PATH_DATA_PROCESSED, "MJR_seq_vst.csv"))
```

---

# Differential Expression Analysis

**Note:** The following chunks perform heavy computation.  
- Set `eval=TRUE` when ready to run the full DEG analysis  
- Results are cached and saved to avoid re-running  

## Prepare for DEG Analysis

```{r prepare-deg-data}
# Filter metadata: remove untreated controls and incomplete donor
seq_degs <- seq_transposed %>%
  filter(EXPOSURE != "Untreated_Control", PATIENTCODE != "DD007T")

# Create proper count matrix (genes as rows, samples as columns)
counts_for_deg <- as.matrix(t(seq_degs[, 1:(ncol(seq_degs) - 3)]))
colnames(counts_for_deg) <- rownames(seq_degs)

# Metadata for DEG analysis
meta_deg <- seq_degs %>%
  select(EXPOSURE, PATIENTCODE, SEX) %>%
  mutate(
    EXPOSURE = factor(EXPOSURE),
    PATIENTCODE = factor(PATIENTCODE),
    SEX = factor(SEX)
  )

# Create DGEList object
dataset_MJR <- DGEList(counts_for_deg, group = meta_deg$EXPOSURE)

cat("✓ Data prepared for DEG analysis\n")
cat("  - Genes:", nrow(counts_for_deg), "\n")
cat("  - Samples:", ncol(counts_for_deg), "\n")
```

## DREAM Analysis: Exposure Effects

```{r deg-exposure, eval=TRUE, cache=TRUE}
# ============================================================================
# EXPOSURE EFFECTS (NO SEX INTERACTION)
# ============================================================================
# Set eval=TRUE to run this analysis (takes ~1 hour)

param_MJR <- SnowParam(DREAM_N_CORES, "SOCK", progressbar = TRUE)

# Define model
model_MJR <- ~ 0 + EXPOSURE + (1 | PATIENTCODE)

# Define contrasts
contrasts_exposure <- c(
  Peat5_vs_PBS = "(EXPOSUREPeat_5 - EXPOSUREPBS_Control)",
  Peat25_vs_PBS = "(EXPOSUREPeat_25 - EXPOSUREPBS_Control)",
  Euc5_vs_PBS = "(EXPOSUREEucalyptus_5 - EXPOSUREPBS_Control)",
  Euc25_vs_PBS = "(EXPOSUREEucalyptus_25 - EXPOSUREPBS_Control)",
  Pine5_vs_PBS = "(EXPOSUREPine_5 - EXPOSUREPBS_Control)",
  Pine25_vs_PBS = "(EXPOSUREPine_25 - EXPOSUREPBS_Control)",
  RO5_vs_PBS = "(EXPOSURERedOak_5 - EXPOSUREPBS_Control)",
  RO25_vs_PBS = "(EXPOSURERedOak_25 - EXPOSUREPBS_Control)",
  Peat25_vs_Euc25 = "(EXPOSUREPeat_25 - EXPOSUREEucalyptus_25)",
  Peat25_vs_Pine25 = "(EXPOSUREPeat_25 - EXPOSUREPine_25)",
  Peat25_vs_RO25 = "(EXPOSUREPeat_25 - EXPOSURERedOak_25)",
  Euc25_vs_Pine25 = "(EXPOSUREEucalyptus_25 - EXPOSUREPine_25)",
  Euc25_vs_RO25 = "(EXPOSUREEucalyptus_25 - EXPOSURERedOak_25)",
  Pine25_vs_RO25 = "(EXPOSUREPine_25 - EXPOSURERedOak_25)"
)

# Create contrast matrix
list_MJR <- makeContrastsDream(model_MJR, meta_deg, contrasts = contrasts_exposure)

# Calculate voom with Dream weights
dream <- voomWithDreamWeights(dataset_MJR, model_MJR, meta_deg, BPPARAM = param_MJR)

# Variance partition analysis
form <- ~ (1 | PATIENTCODE) + (1 | EXPOSURE)
vp <- fitExtractVarPartModel(dream, form, meta_deg)
plotVarPart(sortCols(vp))

# Fit linear model
dream_fit <- dream(dream, model_MJR, meta_deg, list_MJR)
dream_fit <- eBayes(dream_fit)

# Extract DEGs
dream_DEGs <- lapply(names(contrasts_exposure), function(x) {
  top <- topTable(dream_fit, coef = x, sort.by = "P", n = Inf)
  data.frame(
    log2FC = top$logFC,
    P.Value = top$P.Value,
    adj.P.Val = top$adj.P.Val,
    Expression = top$AveExpr,
    row.names = rownames(top)
  )
})
names(dream_DEGs) <- names(contrasts_exposure)

# Save results
saveRDS(list(fit = dream_fit, degs = dream_DEGs, voom = dream, vp = vp),
        here::here(PATH_DATA_PROCESSED, "dream_exposure_results.rds"))

cat("✓ DREAM exposure analysis complete\n")
```

## DREAM Analysis: Sex × Exposure Interaction

```{r deg-sex-exposure, eval=TRUE, cache=TRUE}
# ============================================================================
# SEX × EXPOSURE INTERACTION
# ============================================================================
# Set eval=TRUE to run this analysis (takes ~1 hour)

# Filter out incomplete donor
seq_sex <- seq_degs %>% filter(PATIENTCODE != "DD007T")

# Create interaction variable
seq_sex$EXPOSURE_SEX <- paste(seq_sex$EXPOSURE, seq_sex$SEX, sep = "_")

# Prepare data
counts_sex <- as.matrix(t(seq_sex[, 1:(ncol(seq_sex) - 4)]))
colnames(counts_sex) <- rownames(seq_sex)

meta_sex <- seq_sex %>%
  select(EXPOSURE_SEX, PATIENTCODE, EXPOSURE, SEX) %>%
  mutate(
    EXPOSURE_SEX = factor(EXPOSURE_SEX),
    PATIENTCODE = factor(PATIENTCODE)
  )

dataset_MJR_sex <- DGEList(counts_sex, group = meta_sex$EXPOSURE_SEX)

# Define model
model_MJR_sex <- ~ 0 + EXPOSURE_SEX + (1 | PATIENTCODE)

# Define contrasts
contrasts_sex <- c(
  Peat5_vs_PBS_M = "(EXPOSURE_SEXPeat_5_M - EXPOSURE_SEXPBS_Control_M)",
  Peat25_vs_PBS_M = "(EXPOSURE_SEXPeat_25_M - EXPOSURE_SEXPBS_Control_M)",
  Euc5_vs_PBS_M = "(EXPOSURE_SEXEucalyptus_5_M - EXPOSURE_SEXPBS_Control_M)",
  Euc25_vs_PBS_M = "(EXPOSURE_SEXEucalyptus_25_M - EXPOSURE_SEXPBS_Control_M)",
  Pine5_vs_PBS_M = "(EXPOSURE_SEXPine_5_M - EXPOSURE_SEXPBS_Control_M)",
  Pine25_vs_PBS_M = "(EXPOSURE_SEXPine_25_M - EXPOSURE_SEXPBS_Control_M)",
  RO5_vs_PBS_M = "(EXPOSURE_SEXRedOak_5_M - EXPOSURE_SEXPBS_Control_M)",
  RO25_vs_PBS_M = "(EXPOSURE_SEXRedOak_25_M - EXPOSURE_SEXPBS_Control_M)",
  Peat5_vs_PBS_F = "(EXPOSURE_SEXPeat_5_F - EXPOSURE_SEXPBS_Control_F)",
  Peat25_vs_PBS_F = "(EXPOSURE_SEXPeat_25_F - EXPOSURE_SEXPBS_Control_F)",
  Euc5_vs_PBS_F = "(EXPOSURE_SEXEucalyptus_5_F - EXPOSURE_SEXPBS_Control_F)",
  Euc25_vs_PBS_F = "(EXPOSURE_SEXEucalyptus_25_F - EXPOSURE_SEXPBS_Control_F)",
  Pine5_vs_PBS_F = "(EXPOSURE_SEXPine_5_F - EXPOSURE_SEXPBS_Control_F)",
  Pine25_vs_PBS_F = "(EXPOSURE_SEXPine_25_F - EXPOSURE_SEXPBS_Control_F)",
  RO5_vs_PBS_F = "(EXPOSURE_SEXRedOak_5_F - EXPOSURE_SEXPBS_Control_F)",
  RO25_vs_PBS_F = "(EXPOSURE_SEXRedOak_25_F - EXPOSURE_SEXPBS_Control_F)"
)

list_MJR_sex <- makeContrastsDream(model_MJR_sex, meta_sex, contrasts = contrasts_sex)

# Calculate voom weights
dream_sex <- voomWithDreamWeights(dataset_MJR_sex, model_MJR_sex, meta_sex, BPPARAM = param_MJR)

# Variance partition
form_sex <- ~ (1 | PATIENTCODE) + (1 | EXPOSURE) + (1 | SEX)
vp_sex <- fitExtractVarPartModel(dream_sex, form_sex, meta_sex)
plotVarPart(sortCols(vp_sex))

# Fit model
dream_fit_sex <- dream(dream_sex, model_MJR_sex, meta_sex, list_MJR_sex)
dream_fit_sex <- eBayes(dream_fit_sex)

# Extract DEGs
dream_DEGs_sex <- lapply(names(contrasts_sex), function(x) {
  top <- topTable(dream_fit_sex, coef = x, sort.by = "P", n = Inf)
  data.frame(
    log2FC = top$logFC,
    P.Value = top$P.Value,
    adj.P.Val = top$adj.P.Val,
    Expression = top$AveExpr,
    row.names = rownames(top)
  )
})
names(dream_DEGs_sex) <- names(contrasts_sex)

# Save results
saveRDS(list(fit = dream_fit_sex, degs = dream_DEGs_sex, voom = dream_sex, vp = vp_sex),
        here::here(PATH_DATA_PROCESSED, "dream_sex_results.rds"))

cat("✓ DREAM sex interaction analysis complete\n")
```

## Load Pre-computed Results

```{r load-deg-results}
# Load pre-computed DEG results if they exist
if (file.exists(here::here(PATH_DATA_PROCESSED, "dream_exposure_results.rds"))) {
  dream_results <- readRDS(here::here(PATH_DATA_PROCESSED, "dream_exposure_results.rds"))
  dream_DEGs <- dream_results$degs
  cat("✓ Loaded pre-computed exposure DEG results\n")
  cat("  - Contrasts:", length(dream_DEGs), "\n")
} else {
  cat("⚠ No pre-computed exposure results found. Run deg-exposure chunk with eval=TRUE\n")
}

if (file.exists(here::here(PATH_DATA_PROCESSED, "dream_sex_results.rds"))) {
  dream_sex_results <- readRDS(here::here(PATH_DATA_PROCESSED, "dream_sex_results.rds"))
  dream_DEGs_sex <- dream_sex_results$degs
  cat("✓ Loaded pre-computed sex interaction DEG results\n")
  cat("  - Contrasts:", length(dream_DEGs_sex), "\n")
} else {
  cat("⚠ No pre-computed sex results found. Run deg-sex-exposure chunk with eval=TRUE\n")
}
```

---

# DEG Classification & Summary

## Classify DEGs

```{r classify-degs}
# Combine all DEG results into one data frame
combine_deg_results <- function(deg_list, add_gene_col = TRUE) {
  combined <- lapply(names(deg_list), function(name) {
    df <- deg_list[[name]]
    if (add_gene_col) df$X <- rownames(df)
    df$file_name <- name
    df$sample_name <- sub("_.*", "", name)
    df$Sex <- ifelse(grepl("_F$", name), "Female",
                     ifelse(grepl("_M$", name), "Male", NA))
    return(df)
  })
  bind_rows(combined)
}

# Combine exposure and sex DEGs
if (exists("dream_DEGs")) {
  combined_degs_exposure <- combine_deg_results(dream_DEGs)
}

if (exists("dream_DEGs_sex")) {
  combined_degs_sex <- combine_deg_results(dream_DEGs_sex)
}

# Merge all
if (exists("combined_degs_exposure") && exists("combined_degs_sex")) {
  combined_degs <- bind_rows(combined_degs_exposure, combined_degs_sex)
} else if (exists("combined_degs_exposure")) {
  combined_degs <- combined_degs_exposure
} else {
  cat("⚠ No DEG results loaded\n")
}

# Classify DEGs
if (exists("combined_degs")) {
  combined_degs <- combined_degs %>%
    mutate(
      DEG = case_when(
        log2FC >= LOG2FC_CUTOFF & adj.P.Val < ADJ_P_CUTOFF ~ "UP",
        log2FC <= -LOG2FC_CUTOFF & adj.P.Val < ADJ_P_CUTOFF ~ "DOWN",
        TRUE ~ "NO"
      )
    )
  
  cat("✓ DEG classification complete\n")
  
  # Summary table
  deg_summary <- combined_degs %>%
    filter(DEG != "NO") %>%
    group_by(sample_name, Sex, DEG) %>%
    summarise(n_genes = n(), .groups = "drop") %>%
    pivot_wider(names_from = DEG, values_from = n_genes, values_fill = 0)
  
  print(deg_summary)
}
```

## Export DEG Results

```{r export-degs}
if (exists("combined_degs")) {
  # Export full results
  write_csv(combined_degs, here::here(PATH_OUTPUT_TABLES, "all_deg_results.csv"))
  
  # Export significant DEGs only
  sig_degs <- combined_degs %>% filter(DEG != "NO")
  write_csv(sig_degs, here::here(PATH_OUTPUT_TABLES, "significant_degs.csv"))
  
  # Export by fuel type (high dose only)
  degs_by_fuel <- combined_degs %>%
    filter(is.na(Sex), grepl("25", sample_name), DEG != "NO") %>%
    select(X, log2FC, P.Value, adj.P.Val, Expression, sample_name)
  write_csv(degs_by_fuel, here::here(PATH_OUTPUT_TABLES, "degs_by_fuel.csv"))
  
  # Export by sex (high dose only)
  degs_by_sex <- combined_degs %>%
    filter(!is.na(Sex), grepl("25", sample_name), DEG != "NO") %>%
    select(X, log2FC, P.Value, adj.P.Val, Expression, sample_name, Sex)
  write_csv(degs_by_sex, here::here(PATH_OUTPUT_TABLES, "degs_by_sex.csv"))
  
  cat("✓ DEG results exported\n")
}
```


---

## IPA Export Tables

```{r export-ipa-tables}
if (exists("combined_degs")) {
  
  # ============================================================================
  # IPA Export: FUEL TYPE (no sex interaction)
  # ============================================================================
  
  IPA_degs_fuel <- combined_degs %>% 
    filter(!grepl("(_F|_M)$", file_name)) %>%  # Exclude sex-stratified results
    select(-P.Value, -Expression)  # Keep only log2FC and adj.P.Val
  
  # Pivot to wide format
  IPA_degs_fuel_wide <- IPA_degs_fuel %>%
    pivot_wider(
      names_from = file_name, 
      values_from = c(log2FC, adj.P.Val)
    )
  
  # Filter for high-dose (25) comparisons only
  IPA_degs_fuel_filtered <- IPA_degs_fuel_wide %>% 
    select(X, contains("25"))
  
  # Save
  save_table(IPA_degs_fuel_filtered, "IPA_DEGs_FUEL.csv")
  
  cat("✓ IPA fuel type table exported\n")
  cat("  - Genes:", nrow(IPA_degs_fuel_filtered), "\n")
  cat("  - Comparisons:", (ncol(IPA_degs_fuel_filtered) - 1) / 2, "\n")
  
  # ============================================================================
  # IPA Export: SEX-STRATIFIED
  # ============================================================================
  
  IPA_degs_sex <- combined_degs %>% 
    filter(grepl("(_F|_M)$", file_name)) %>%  # Only sex-stratified results
    select(-P.Value, -Expression)
  
  # Pivot to wide format
  IPA_degs_sex_wide <- IPA_degs_sex %>%
    pivot_wider(
      names_from = file_name, 
      values_from = c(log2FC, adj.P.Val)
    )
  
  # Filter for high-dose (25) comparisons only
  IPA_degs_sex_filtered <- IPA_degs_sex_wide %>% 
    select(X, contains("25"))
  
  # Save
  save_table(IPA_degs_sex_filtered, "IPA_DEGs_SEX.csv")
  
  cat("✓ IPA sex-stratified table exported\n")
  cat("  - Genes:", nrow(IPA_degs_sex_filtered), "\n")
  cat("  - Comparisons:", (ncol(IPA_degs_sex_filtered) - 1) / 2, "\n")
  
} else {
  cat("⚠ combined_degs not found. Run DEG analysis first.\n")
}
```

---

# Visualization

## Volcano Plots: Fuel Type

```{r volcano-fuel, fig.width=14, fig.height=8}
if (exists("combined_degs")) {
  # Filter for high-dose fuel comparisons
  fuel_degs <- combined_degs %>%
    filter(is.na(Sex), grepl("25_vs_PBS", file_name)) %>%
    mutate(sample_name = case_when(
      sample_name == "Euc25" ~ "Eucalyptus 25 µg/cm²",
      sample_name == "Peat25" ~ "Peat 25 µg/cm²",
      sample_name == "Pine25" ~ "Pine 25 µg/cm²",
      sample_name == "RO25" ~ "Red Oak 25 µg/cm²",
      TRUE ~ sample_name
    ))
  
  # Define highlighted genes
  selected_genes <- c('SPDEF', 'IL1A', 'IL1B', 'SRXN1')
  
  # Add labels for top genes
  fuel_degs <- fuel_degs %>%
    group_by(file_name, DEG) %>%
    mutate(
      delabel = case_when(
        DEG == "UP" & row_number(adj.P.Val) <= 6 ~ X,
        DEG == "DOWN" & row_number(adj.P.Val) <= 6 ~ X,
        TRUE ~ NA_character_
      )
    ) %>%
    ungroup()
  
  # Count UP and DOWN genes for each sample_name
  up_counts <- fuel_degs %>%
    filter(DEG == "UP") %>%
    group_by(sample_name) %>%
    summarise(n = n(), .groups = "drop")
  
  down_counts <- fuel_degs %>%
    filter(DEG == "DOWN") %>%
    group_by(sample_name) %>%
    summarise(n = n(), .groups = "drop")
  
  # Create volcano plot
  p_volcano_fuel <- ggplot(data = fuel_degs, aes(x = log2FC, y = -log10(adj.P.Val), color = DEG, label = delabel)) +
    geom_vline(xintercept = c(-1, 1), col = "gray", linetype = 'dashed') +
    geom_hline(yintercept = -log10(0.05), col = "gray", linetype = 'dashed') +
    
    # Plot all genes first
    geom_point(alpha = 0.7, size = 2) +
    
    # Overlay labeled genes with black outlines
    geom_point(data = subset(fuel_degs, !is.na(delabel) | X %in% selected_genes),
               aes(x = log2FC, y = -log10(adj.P.Val)),
               color = "black", size = 2, shape = 21, stroke = 1) +
    
    # Add labels with white boxes
    geom_label_repel(data = subset(fuel_degs, !is.na(delabel) | X %in% selected_genes),
                     aes(label = ifelse(!is.na(delabel), delabel, X)),
                     size = 3, fontface = "bold", fill = "white", color = "black",
                     box.padding = 0.4, segment.color = "black", max.overlaps = 20) +
    
    # Annotate UP count at the top-right corner
    geom_text(data = up_counts, aes(x = 3.5, y = 30.5, label = paste0("UP: ", n)),
              color = "#bb0c00", size = 3, fontface = "bold", hjust = 1, inherit.aes = FALSE) +
    
    # Annotate DOWN count at the top-left corner
    geom_text(data = down_counts, aes(x = -4.5, y = 30.5, label = paste0("DOWN: ", n)),
              color = "#00AFBB", size = 3, fontface = "bold", hjust = 0, inherit.aes = FALSE) +
    
    scale_color_manual(
      values = c("DOWN" = "#00AFBB", "NO" = "grey", "UP" = "#bb0c00"),
      labels = c("Downregulated", "Not significant", "Upregulated")
    ) +
    coord_cartesian(ylim = c(0, 30), xlim = c(-8, 8)) +
    labs(
      x = expression("log"[2] * "FC"),
      y = expression("-log"[10] * "adj.P.Val")
    ) +
    theme_minimal(base_size = 14) +
    theme(
      strip.text = element_text(size = 12, face = "bold"),
      plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
      panel.border = element_rect(color = "black", fill = NA, linewidth = .5),
      panel.background = element_rect(fill = "white", color = NA),
      plot.background = element_rect(fill = "#ffffff", color = NA),
      legend.background = element_rect(fill = "#ffffff", color = NA),
      legend.key = element_rect(fill = "#ffffff", color = NA),
      panel.grid.major = element_line(color = "gray80"),
      panel.grid.minor = element_line(color = "gray90")
    ) +
    facet_wrap(~ sample_name)
  
  print(p_volcano_fuel)
  save_plot("volcano_plot_fuel_type.png", p_volcano_fuel, width = 14, height = 8, bg = "white")
}
```

## Volcano Plots: Sex Comparison

```{r volcano-sex, fig.width=14, fig.height=10}
if (exists("combined_degs")) {
  # Filter for high-dose sex comparisons
  sex_degs_high <- combined_degs %>%
    filter(!is.na(Sex), grepl("25_vs_PBS", file_name)) %>%
    mutate(sample_name = case_when(
      sample_name == "Euc25" ~ "Eucalyptus 25 µg/cm²",
      sample_name == "Peat25" ~ "Peat 25 µg/cm²",
      sample_name == "Pine25" ~ "Pine 25 µg/cm²",
      sample_name == "RO25" ~ "Red Oak 25 µg/cm²",
      TRUE ~ sample_name
    ))
  
  # Add labels
  sex_degs_high <- sex_degs_high %>%
    group_by(file_name, DEG) %>%
    mutate(
      delabel = case_when(
        DEG == "UP" & row_number(adj.P.Val) <= 6 ~ X,
        DEG == "DOWN" & row_number(adj.P.Val) <= 6 ~ X,
        X %in% highlighted_genes ~ X,
        TRUE ~ NA_character_
      )
    ) %>%
    ungroup()
  
  # Define highlighted genes
  selected_genes <- c('SPDEF', 'IL1A', 'IL1B', 'SRXN1')
  
  # Function to plot sex volcano with background color
  plot_sex_degs <- function(sex_choice, selected_genes) {     
    # Ensure Male is on top in the facet grid   
    sex_degs_high$Sex <- factor(sex_degs_high$Sex, levels = c("Male", "Female"))    
    
    # Filter data based on the specified sex     
    filtered_data <- subset(sex_degs_high, Sex == sex_choice)    
    
    # Specify the background color for the sex chosen     
    background_color <- ifelse(sex_choice == "Female", "#fff0f3", "#F0FFF0")    
    
    # Count UP and DOWN genes for each sample_name
    up_counts <- filtered_data %>%
      dplyr::filter(DEG == "UP") %>%
      dplyr::group_by(Sex, sample_name) %>%
      dplyr::summarise(n = n(), .groups = "drop")
    
    down_counts <- filtered_data %>%
      dplyr::filter(DEG == "DOWN") %>%
      dplyr::group_by(Sex, sample_name) %>%
      dplyr::summarise(n = n(), .groups = "drop")
    
    # Generate the plot     
    p <- ggplot(data = filtered_data, aes(x = log2FC, y = -log10(adj.P.Val), color = DEG, label = delabel)) +       
      geom_vline(xintercept = c(-1, 1), col = "gray", linetype = 'dashed') +       
      geom_hline(yintercept = -log10(0.05), col = "gray", linetype = 'dashed') +      
      geom_point(alpha = 0.7, size = 2) +      
      geom_point(data = subset(filtered_data, !is.na(delabel) | X %in% selected_genes),                  
                 aes(x = log2FC, y = -log10(adj.P.Val)),                  
                 color = "black", size = 2, shape = 21, stroke = 1) +      
      geom_label_repel(data = subset(filtered_data, !is.na(delabel) | X %in% selected_genes),                        
                       aes(label = ifelse(!is.na(delabel), delabel, X)),                        
                       size = 3, fontface = "bold", fill = "white", color = "black",                        
                       box.padding = 0.4, segment.color = "black", max.overlaps = 20) +      
      geom_text(data = up_counts, aes(x = 3.5, y = 20.5, label = paste0("UP: ", n)), 
                color = "#bb0c00", size = 3, fontface = "bold", hjust = 1, inherit.aes = FALSE) +
      geom_text(data = down_counts, aes(x = -4.5, y = 20.5, label = paste0("DOWN: ", n)), 
                color = "#00AFBB", size = 3, fontface = "bold", hjust = 0, inherit.aes = FALSE) +
      scale_color_manual(           
        values = c("DOWN" = "#00AFBB", "NO" = "grey", "UP" = "#bb0c00"),           
        labels = c("Downregulated", "Not significant", "Upregulated")       
      ) +       
      coord_cartesian(ylim = c(0, 20), xlim = c(-8, 8)) +       
      labs(           
        x = expression("log"[2] * "FC"),           
        y = expression("-log"[10] * "adj.P.Val")       
      ) +       
      theme_minimal(base_size = 14) +         
      theme(           
        strip.text = element_text(size = 12, face = "bold"),           
        plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),           
        panel.border = element_rect(color = "black", fill = NA, linewidth = .5),         
        panel.background = element_rect(fill = background_color, color = NA),  # BACKGROUND COLOR HERE
        plot.background = element_rect(fill = "#ffffff", color = NA),         
        legend.background = element_rect(fill = "#ffffff", color = NA),         
        legend.key = element_rect(fill = "#ffffff", color = NA),
        panel.grid.major = element_line(color = "gray80"),         
        panel.grid.minor = element_line(color = "gray90")
      ) +          
      facet_grid(Sex ~ sample_name, scales = "free_y")
    
    return(p)
  }
  
  # Create plots for both sexes
  p_male <- plot_sex_degs("Male", selected_genes)
  p_female <- plot_sex_degs("Female", selected_genes)
  
  print(p_male)
  print(p_female)
  
  save_plot("volcano_plot_male.png", p_male, width = 16, height = 4)
  save_plot("volcano_plot_female.png", p_female, width = 16, height = 4)
}
```


## UpSet Plots: DEG Overlap

```{r upset-plots, fig.width=12, fig.height=8}
library(ComplexUpset)

if (exists("combined_degs")) {
  # Custom fuel colors
  fuel_colors <- c(
    "Red Oak 25 µg/cm²" = "brown3",
    "Pine 25 µg/cm²" = "deepskyblue3",
    "Peat 25 µg/cm²" = "gold3",
    "Eucalyptus 25 µg/cm²" = "darkorchid"
  )
  
  # ============================================================================
  # PLOT 1: Fuel type UpSet (all samples combined)
  # ============================================================================
  
  # Prepare data
  degs_by_fuel_upset <- combined_degs %>%
    filter(is.na(Sex), grepl("25_vs_PBS", file_name), DEG != "NO") %>%
    mutate(sample_name = case_when(
      sample_name == "Euc25" ~ "Eucalyptus 25 µg/cm²",
      sample_name == "Peat25" ~ "Peat 25 µg/cm²",
      sample_name == "Pine25" ~ "Pine 25 µg/cm²",
      sample_name == "RO25" ~ "Red Oak 25 µg/cm²",
      TRUE ~ sample_name
    )) %>%
    distinct(X, sample_name)
  
  if (nrow(degs_by_fuel_upset) > 0) {
    deg_wide <- degs_by_fuel_upset %>%
      mutate(present = TRUE) %>%
      pivot_wider(
        names_from = sample_name,
        values_from = present,
        values_fill = FALSE
      )
    
    fuel_sets <- names(fuel_colors)
    
    fuel_plot <- upset(
      deg_wide,
      intersect = fuel_sets,
      set_sizes = upset_set_size(
        geom = geom_bar(aes(fill = group), width = 0.6)
      ),
      queries = list(
        upset_query(set = "Red Oak 25 µg/cm²", color = "brown3", fill = "brown3"),
        upset_query(set = "Pine 25 µg/cm²", color = "deepskyblue3", fill = "deepskyblue3"),
        upset_query(set = "Peat 25 µg/cm²", color = "gold3", fill = "gold3"),
        upset_query(set = "Eucalyptus 25 µg/cm²", color = "darkorchid", fill = "darkorchid")
      ),
      base_annotations = list(
        'Intersection size' = intersection_size(
          counts = TRUE,
          bar_number_threshold = 1,
          text = list(size = 3)
        )
      ),
      matrix = (
        intersection_matrix(
          geom = geom_point(aes(color = group), size = 3, shape = 19),
          segment = geom_segment(color = "black", size = 0.5),
          outline_color = list(active = "black", inactive = "grey70")
        )
        + scale_color_manual(values = fuel_colors, breaks = fuel_sets)
      ),
      themes = upset_default_themes(text = element_text(size = 12))
    ) +
      ggtitle("UpSet Plot of DEGs by Fuel Type") +
      theme(
        plot.title = element_text(hjust = 0.5, size = 14, face = "bold")
      )
    
    print(fuel_plot)
    save_plot("DEGs_by_FuelType_UpSet.png", fuel_plot, width = 12, height = 8, bg = "white")
    
    # ============================================================================
    # PLOT 2 & 3: Sex-stratified UpSet plots
    # ============================================================================
    
    # Prepare sex-stratified data
    degs_by_sex_upset <- combined_degs %>%
      filter(!is.na(Sex), grepl("25_vs_PBS", file_name), DEG != "NO") %>%
      mutate(sample_name = case_when(
        sample_name == "Euc25" ~ "Eucalyptus 25 µg/cm²",
        sample_name == "Peat25" ~ "Peat 25 µg/cm²",
        sample_name == "Pine25" ~ "Pine 25 µg/cm²",
        sample_name == "RO25" ~ "Red Oak 25 µg/cm²",
        TRUE ~ sample_name
      ))
    
    # Male data
    male_data <- degs_by_sex_upset %>%
      filter(Sex == "Male") %>%
      distinct(X, sample_name) %>%
      mutate(present = TRUE) %>%
      pivot_wider(
        names_from = sample_name,
        values_from = present,
        values_fill = FALSE
      )
    
    # Female data
    female_data <- degs_by_sex_upset %>%
      filter(Sex == "Female") %>%
      distinct(X, sample_name) %>%
      mutate(present = TRUE) %>%
      pivot_wider(
        names_from = sample_name,
        values_from = present,
        values_fill = FALSE
      )
    
    # Function to create upset plot with proper colors
    create_upset_plot <- function(data, title_text) {
      upset(
        data,
        intersect = fuel_sets,
        set_sizes = upset_set_size(
          geom = geom_bar(aes(fill = group), width = 0.6)
        ),
        queries = list(
          upset_query(set = "Red Oak 25 µg/cm²", color = "brown3", fill = "brown3"),
          upset_query(set = "Pine 25 µg/cm²", color = "deepskyblue3", fill = "deepskyblue3"),
          upset_query(set = "Peat 25 µg/cm²", color = "gold3", fill = "gold3"),
          upset_query(set = "Eucalyptus 25 µg/cm²", color = "darkorchid", fill = "darkorchid")
        ),
        base_annotations = list(
          'Intersection size' = intersection_size(
            counts = TRUE,
            bar_number_threshold = 1,
            text = list(size = 3)
          )
        ),
        matrix = (
          intersection_matrix(
            geom = geom_point(aes(color = group), size = 3, shape = 19),
            segment = geom_segment(color = "black", size = 0.5),
            outline_color = list(active = "black", inactive = "grey70")
          )
          + scale_color_manual(values = fuel_colors, breaks = fuel_sets)
        ),
        themes = upset_default_themes(text = element_text(size = 12))
      ) +
        ggtitle(title_text) +
        theme(plot.title = element_text(hjust = 0.5, size = 14, face = "bold"))
    }
    
    # Create and save male plot
    if (nrow(male_data) > 1) {
      male_plot <- create_upset_plot(male_data, "DEGs by Fuel Type - Males")
      print(male_plot)
      save_plot("DEGs_Male_UpSet.png", male_plot, width = 12, height = 8, bg = "white")
    }
    
    # Create and save female plot
    if (nrow(female_data) > 1) {
      female_plot <- create_upset_plot(female_data, "DEGs by Fuel Type - Females")
      print(female_plot)
      save_plot("DEGs_Female_UpSet.png", female_plot, width = 12, height = 8, bg = "white")
    }
    
    cat("\n✓ All UpSet plots created and saved\n")
    cat("  - Fuel type combined\n")
    cat("  - Male stratified\n")
    cat("  - Female stratified\n")
  }
}
```

## Euler Diagrams: DEG Overlap

```{r euler-deg-overlap, fig.width=10, fig.height=8}
library(eulerr)

if (exists("combined_degs")) {
  
  # Define custom colors for fuel types
  fuel_colors_euler <- c(
    "RO25" = "brown3", 
    "Pine25" = "deepskyblue3", 
    "Euc25" = "darkorchid", 
    "Peat25" = "gold3"
  )
  
  # ============================================================================
  # FUEL TYPE: UP and DOWN regulated genes
  # ============================================================================
  
  only_fueltype_degs <- combined_degs %>%
    filter(is.na(Sex), grepl("25", sample_name), DEG != "NO") %>%
    mutate(delabel = X)
  
  cat("\n=== FUEL TYPE EULER PLOTS ===\n")
  
  # Loop through UP and DOWN
  for (direction in c("UP", "DOWN")) {
    
    # Get list of genes for each fuel type
    fuel_types <- unique(only_fueltype_degs$sample_name)
    
    eul_list <- lapply(fuel_types, function(fuel) {
      subset(only_fueltype_degs, sample_name == fuel & DEG == direction)$delabel
    })
    
    # Remove duplicates
    eul_list_unique <- lapply(eul_list, unique)
    names(eul_list_unique) <- fuel_types
    
    # Get conserved genes
    overlap_genes <- Reduce(intersect, eul_list_unique)
    cat("\n", direction, "-regulated genes conserved across fuel types:", length(overlap_genes), "\n")
    
    # Set colors
    col <- fuel_colors_euler[fuel_types]
    
    # Fit Euler diagram
    fit <- euler(eul_list_unique)
    
    # Save plot using grDevices
    filename <- here::here(PATH_OUTPUT_FIGURES, paste0("euler_fuel_", direction, ".png"))
    png(filename, width = 10, height = 8, units = "in", res = 300, bg = "white")
    
    # Use print() to force rendering
    print(
      plot(fit,
           fills = list(fill = col, alpha = 0.5),
           legend = list(side = "bottom", col = col, text.col = "black"),
           quantities = list(cex = 1.2, fontface = "bold"),
           main = paste0(direction, "-regulated Genes by Fuel Type"))
    )
    
    dev.off()
    
    cat("✓ Saved:", basename(filename), "\n")
  }
  
  # ============================================================================
  # SEX-STRATIFIED: UP and DOWN regulated genes
  # ============================================================================
  
  only_sex_degs <- combined_degs %>%
    filter(!is.na(Sex), grepl("25", sample_name), DEG != "NO") %>%
    mutate(delabel = X)
  
  cat("\n=== SEX-STRATIFIED EULER PLOTS ===\n")
  
  # Loop through Female and Male
  for (sex in c("Female", "Male")) {
    
    filtered_data <- subset(only_sex_degs, Sex == sex)
    
    # Loop through UP and DOWN
    for (direction in c("UP", "DOWN")) {
      
      filtered_direction_data <- subset(filtered_data, DEG == direction)
      
      fuel_types <- unique(filtered_direction_data$sample_name)
      
      # Skip if no fuel types
      if (length(fuel_types) == 0) {
        cat("⚠ No fuel types for", sex, direction, "\n")
        next
      }
      
      eul_list <- lapply(fuel_types, function(fuel) {
        subset(filtered_direction_data, sample_name == fuel)$delabel
      })
      
      # Remove duplicates
      eul_list_unique <- lapply(eul_list, unique)
      names(eul_list_unique) <- fuel_types
      
      # Get conserved genes
      overlap_genes <- Reduce(intersect, eul_list_unique)
      cat("\n", sex, "-", direction, "genes conserved:", length(overlap_genes), "\n")
      
      # Set colors
      col <- fuel_colors_euler[fuel_types]
      
      # Fit Euler diagram
      fit <- euler(eul_list_unique)
      
      # Save plot
      filename <- here::here(PATH_OUTPUT_FIGURES, paste0("euler_", tolower(sex), "_", direction, ".png"))
      png(filename, width = 10, height = 8, units = "in", res = 300, bg = "white")
      
      # Use print() to force rendering
      print(
        plot(fit,
             fills = list(fill = col, alpha = 0.5),
             legend = list(side = "bottom", col = col, text.col = "black"),
             quantities = list(cex = 1.2, fontface = "bold"),
             main = paste0(sex, " - ", direction, "-regulated Genes by Fuel Type"))
      )
      
      dev.off()
      
      cat("✓ Saved:", basename(filename), "\n")
    }
  }
  
  cat("\n✓ All Euler diagrams saved to:", PATH_OUTPUT_FIGURES, "\n")
  
} else {
  cat("⚠ combined_degs not found. Run DEG analysis first.\n")
}
```

## Gene Expression Barplots

```{r gene-expression-bars, fig.width=12, fig.height=4}
if (exists("vst_counts_df")) {
  # Select genes of interest
  genes_of_interest <- c("IL1A", "SPDEF", "CYP1A1", "TXNRD1")
  
  # Plot gene expression
  p_gene_expr <- plot_gene_expression_bars(
    vst_data = vst_counts_df,
    metadata = meta_data_subset,
    genes = genes_of_interest
  )
  
  print(p_gene_expr)
  save_plot("gene_expression_barplots.png", p_gene_expr, width = 12, height = 4)
}
```

---

# GSEA Pathway Analysis

## Load GSEA Results

```{r load-gsea}
# Define folder containing GSEA HTML results
folder_path_gsea <- here::here("data", "raw", "GSEA")

if (dir.exists(folder_path_gsea)) {
  html_files <- list.files(path = folder_path_gsea, pattern = "\\.html$", full.names = TRUE)
  
  if (length(html_files) > 0) {
    cat("Found", length(html_files), "GSEA HTML files\n")
    
    # Read and parse HTML files
    html_content <- lapply(html_files, function(file) {
      tryCatch(read_html(file), error = function(e) NULL)
    })
    names(html_content) <- tools::file_path_sans_ext(basename(html_files))
    
    # Remove NULL entries (failed reads)
    html_content <- html_content[!sapply(html_content, is.null)]
    
    # Extract tables
    gsea_tables <- lapply(html_content, function(html) {
      table <- html %>% html_element("table")
      if (!is.null(table)) html_table(table) else NULL
    })
    
    # Combine into single data frame
    gsea_results <- bind_rows(
      lapply(names(gsea_tables), function(name) {
        if (!is.null(gsea_tables[[name]])) {
          df <- gsea_tables[[name]]
          df$File_Name <- name
          return(df)
        }
      })
    )
    
    # Clean column names (adjust based on your GSEA output format)
    if (ncol(gsea_results) >= 11) {
      colnames(gsea_results) <- c("Number", "Pathway", "Details", "Size", "ES", "NES", 
                                   "NOM_pval", "FDR_qval", "FWER_pval", "RANK_AT_MAX", 
                                   "LEADING_EDGE", "File_Name")
    }
    
    # Add metadata
    gsea_results <- gsea_results %>%
      mutate(
        sample_name = sub("_.*", "", File_Name),
        Sex = ifelse(grepl("_F_", File_Name), "Female",
                     ifelse(grepl("_M_", File_Name), "Male", NA)),
        FDR_qval = as.numeric(FDR_qval),
        NES = as.numeric(NES)
      )
    
    cat("✓ GSEA results loaded\n")
    cat("  - Total pathways:", nrow(gsea_results), "\n")
    cat("  - Unique comparisons:", length(unique(gsea_results$File_Name)), "\n")
    
  } else {
    cat("⚠ No HTML files found in GSEA folder\n")
  }
} else {
  cat("⚠ GSEA results folder not found at:", folder_path_gsea, "\n")
  cat("  Please check the path or set eval=FALSE for GSEA chunks\n")
}
```

## GSEA Pathway Barplots

```{r gsea-barplots, fig.width=14, fig.height=8}
if (exists("gsea_results")) {
  # Filter for significant pathways
  gsea_sig <- gsea_results %>%
    filter(FDR_qval < FDR_THRESHOLD, !grepl("_(F|M)_", File_Name))
  
  # Plot top pathways by fuel type
  p_gsea_fuel <- plot_gsea_barplot(
    gsea_data = gsea_sig,
    n_top = 12,
    facet_by = "sample_name"
  )
  
  print(p_gsea_fuel)
  save_plot("gsea_pathways_fuel.png", p_gsea_fuel, width = 14, height = 8)
}
```

```{r gsea-barplots-sex, fig.width=14, fig.height=8}
if (exists("gsea_results")) {
  # Filter for sex-stratified, significant pathways
  # DON'T filter by '25' since sample names don't contain it
  gsea_sex_sig <- gsea_results %>%
    filter(!is.na(Sex), FDR_qval < 0.251)  # Removed the grepl("25") filter
  
  cat("After filtering (FDR < 0.251, Sex stratified):", nrow(gsea_sex_sig), "rows\n")
  
  if (nrow(gsea_sex_sig) > 0) {
    cat("Sample breakdown:\n")
    print(table(gsea_sex_sig$sample_name, gsea_sex_sig$Sex))
    
    # Select top pathways and prioritize important ones
    gsea_sex_plot <- gsea_sex_sig %>%
      mutate(
        priority = case_when(
          Pathway %in% paste0("HALLMARK_", priority_pathways) ~ 1,
          TRUE ~ 2
        )
      ) %>%
      group_by(Sex, sample_name) %>%
      arrange(priority, FDR_qval) %>%
      slice_head(n = 12) %>%
      ungroup() %>%
      mutate(
        Path = gsub("^HALLMARK_", "", Pathway),
        Path = gsub("_", " ", Path),
        sample_name_clean = case_when(
          sample_name == "Euc" ~ "Eucalyptus",
          sample_name == "Peat" ~ "Peat",
          sample_name == "Pine" ~ "Pine",
          sample_name %in% c("RO", "RedOak") ~ "Red Oak",
          TRUE ~ sample_name
        )
      )
    
    cat("After selecting top 12 per group:", nrow(gsea_sex_plot), "rows\n")
    cat("Male pathways:", sum(gsea_sex_plot$Sex == "Male"), "\n")
    cat("Female pathways:", sum(gsea_sex_plot$Sex == "Female"), "\n")
    
    # ============================================================================
    # MALE BARPLOT
    # ============================================================================
    male_data <- gsea_sex_plot %>% filter(Sex == "Male")
    
    if (nrow(male_data) > 0) {
      cat("Creating male barplot with", nrow(male_data), "pathways\n")
      
      p_male <- ggplot(male_data, aes(x = reorder(Path, NES), y = NES, fill = FDR_qval)) +
        geom_col() +
        scale_fill_gradient(low = "darkgreen", high = "gray", limits = c(0, 0.35)) +
        geom_hline(yintercept = 0, color = "black", size = 0.5) +
        theme_minimal() +
        theme(
          axis.text.x = element_text(face = "bold", angle = 0, vjust = 0.5, hjust = 0.5),
          axis.text.y = element_text(face = "bold"),
          axis.title.x = element_text(face = "bold"),
          axis.title.y = element_blank(),
          plot.title = element_text(face = "bold", hjust = 0.5),
          legend.title = element_text(face = "bold"),
          legend.text = element_text(face = "bold"),
          panel.grid.major.x = element_line(size = 1),
          axis.line.y = element_line(color = "black", size = 0.5),
          panel.border = element_rect(color = "black", size = 1, fill = NA),
          panel.background = element_blank(),
          strip.text = element_text(size = 12, face = "bold"),
          panel.grid.major.y = element_blank(),
          panel.grid.minor.x = element_blank()
        ) +
        coord_flip() +
        scale_y_continuous(
          limits = c(-2.5, 2.5),
          breaks = seq(-2, 2, by = 1),
          expand = c(0, 0)
        ) +
        labs(
          x = "Pathway",
          y = "Normalized Enrichment Score",
          title = "Male Hallmark Pathways Enrichment Score from GSEA",
          fill = "FDR qval"
        ) +
        facet_wrap(~ sample_name_clean, scales = "free_y")
      
      print(p_male)
      save_plot("Male_Pathways.jpg", p_male, width = 14, height = 8)
      cat("✓ Male barplot saved\n")
    } else {
      cat("⚠ No male pathways to plot\n")
    }
    
    # ============================================================================
    # FEMALE BARPLOT
    # ============================================================================
    female_data <- gsea_sex_plot %>% filter(Sex == "Female")
    
    if (nrow(female_data) > 0) {
      cat("Creating female barplot with", nrow(female_data), "pathways\n")
      
      p_female <- ggplot(female_data, aes(x = reorder(Path, NES), y = NES, fill = FDR_qval)) +
        geom_col() +
        scale_fill_gradient(low = "darkmagenta", high = "gray", limits = c(0, 0.35)) +
        geom_hline(yintercept = 0, color = "black", size = 0.5) +
        theme_minimal() +
        theme(
          axis.text.x = element_text(face = "bold", angle = 0, vjust = 0.5, hjust = 0.5),
          axis.text.y = element_text(face = "bold"),
          axis.title.x = element_text(face = "bold"),
          axis.title.y = element_blank(),
          plot.title = element_text(face = "bold", hjust = 0.5),
          legend.title = element_text(face = "bold"),
          legend.text = element_text(face = "bold"),
          panel.grid.major.x = element_line(size = 1),
          axis.line.y = element_line(color = "black", size = 0.5),
          panel.border = element_rect(color = "black", size = 1, fill = NA),
          panel.background = element_blank(),
          strip.text = element_text(size = 12, face = "bold"),
          panel.grid.major.y = element_blank(),
          panel.grid.minor.x = element_blank()
        ) +
        coord_flip() +
        scale_y_continuous(
          limits = c(-2.5, 2.5),
          breaks = seq(-2, 2, by = 1),
          expand = c(0, 0)
        ) +
        labs(
          x = "Pathway",
          y = "Normalized Enrichment Score",
          title = "Female Hallmark Pathways Enrichment Score from GSEA",
          fill = "FDR qval"
        ) +
        facet_wrap(~ sample_name_clean, scales = "free_y")
      
      print(p_female)
      save_plot("Female_Pathways.jpg", p_female, width = 14, height = 8)
      cat("✓ Female barplot saved\n")
    } else {
      cat("⚠ No female pathways to plot\n")
    }
    
    cat("\n✓ Sex-stratified GSEA barplots complete\n")
  } else {
    cat("⚠ No significant sex-stratified GSEA results found\n")
  }
}
```

## GSEA Dot Plot

```{r gsea-dotplot, fig.width=12, fig.height=8}
if (exists("gsea_results")) {
  # Define pathway list for dot plots
  pathway_list <- c("REACTIVE_OXYGEN_SPECIES_PATHWAY", "ANDROGEN_RESPONSE", 
                    "ESTROGEN_RESPONSE_EARLY", "IL6_JAK_STAT3_SIGNALING", 
                    "TNFA_SIGNALING_VIA_NFKB", "TGF_BETA_SIGNALING", 
                    "OXIDATIVE_PHOSPHORYLATION", "G2M_CHECKPOINT", "GLYCOLYSIS")
  
  # ============================================================================
  # PREPARE DATA
  # ============================================================================
  
  # Sex-stratified data
  sex_df_relevant <- gsea_results %>%
    filter(!is.na(Sex)) %>%
    mutate(
      clean_path = gsub("HALLMARK_", "", Pathway),
      clean_path = gsub("_", " ", clean_path),
      sample_sex = paste(sample_name, Sex, sep = "_")
    )
  
  # Fuel type only data
  fueltype_df_relevant <- gsea_results %>%
    filter(is.na(Sex)) %>%
    mutate(
      clean_path = gsub("HALLMARK_", "", Pathway),
      clean_path = gsub("_", " ", clean_path)
    )
  
  # Filter for significant and relevant pathways
  sex_df_relevant_subset <- sex_df_relevant %>%
    filter(FDR_qval < 0.251) %>%
    filter(clean_path %in% gsub("_", " ", pathway_list))
  
  fueltype_df_relevant_subset <- fueltype_df_relevant %>%
    filter(FDR_qval < 0.251) %>%
    filter(clean_path %in% gsub("_", " ", pathway_list))
  
  # ============================================================================
  # PLOT 1: SEX-STRATIFIED DOT PLOT
  # ============================================================================
  
  if (nrow(sex_df_relevant_subset) > 0) {
    # CORRECT ordering: sort by fuel type (before underscore), NOT by sex
    # This keeps F and M together for each fuel type
    all_combos <- unique(sex_df_relevant_subset$sample_sex)
    
    # Sort by fuel type first (everything before the underscore)
    ordered_combos <- all_combos[order(sub("_.*", "", all_combos), sub(".*_", "", all_combos))]
    
    sex_df_relevant_subset$sample_sex <- factor(
      sex_df_relevant_subset$sample_sex,
      levels = ordered_combos
    )
    
    cat("X-axis order:", paste(levels(sex_df_relevant_subset$sample_sex), collapse = ", "), "\n")
    
    p89 <- ggplot(sex_df_relevant_subset, 
                  aes(x = sample_sex, y = clean_path, size = abs(NES), color = NES)) + 
      geom_point() + 
      scale_size_continuous(range = c(2, 10), limits = c(1, 2.4)) +
      scale_color_gradient2(
        low = "blue", mid = "white", high = "red", midpoint = 0,
        limits = c(-2.3, 2.3),
        name = "NES"
      ) + 
      theme_minimal() + 
      theme(
        axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
        axis.text.y = element_text(face = "bold", hjust = 1),
        axis.title.x = element_text(face = "bold"),
        axis.title.y = element_text(face = "bold"),
        plot.title = element_text(face = "bold"),
        legend.title = element_text(face = "bold"),
        legend.text = element_text(face = "bold"),
        legend.key = element_rect(fill = "white"),
        panel.grid.major = element_line(color = "black"),
        panel.grid.minor = element_line(color = "black")
      ) + 
      labs(
        title = "GSEA Results by Fuel and Sex",
        x = "Fuel Type_Sex",
        y = "Pathway",
        size = "Absolute NES",
        color = "NES"
      ) + 
      guides(color = guide_colorbar(title = "NES", title.position = "top"))
    
    print(p89)
    save_plot("Sex_Dot_Plot.png", p89, width = 8, height = 6, bg = "white")
    
    cat("✓ Sex dot plot created\n")
  } else {
    cat("⚠ No significant sex-stratified pathways\n")
  }
  
  # ============================================================================
  # PLOT 2: FUEL TYPE DOT PLOT
  # ============================================================================
  
  if (nrow(fueltype_df_relevant_subset) > 0) {
    p90 <- ggplot(fueltype_df_relevant_subset, 
                  aes(x = sample_name, y = clean_path, size = abs(NES), color = NES)) + 
      geom_point() + 
      scale_size_continuous(range = c(2, 10), limits = c(1, 2.4)) +
      scale_color_gradient2(
        low = "blue", mid = "white", high = "red", midpoint = 0,
        limits = c(-2.3, 2.3),
        name = "NES"
      ) + 
      theme_minimal() + 
      theme(
        axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
        axis.text.y = element_text(face = "bold"),
        axis.title.x = element_text(face = "bold"),
        axis.title.y = element_text(face = "bold"),
        plot.title = element_text(face = "bold"),
        legend.title = element_text(face = "bold"),
        legend.text = element_text(face = "bold"),
        legend.key = element_rect(fill = "white"),
        panel.grid.major = element_line(color = "black"),
        panel.grid.minor = element_line(color = "black")
      ) + 
      labs(
        title = "GSEA Results by Fuel Type",
        x = "Fuel Type",
        y = "Pathway",
        size = "Absolute NES",
        color = "NES"
      ) + 
      guides(color = guide_colorbar(title = "NES", title.position = "top"))
    
    print(p90)
    save_plot("Fuel_Dot_Plot.png", p90, width = 8, height = 5, bg = "white")
    
    cat("✓ Fuel dot plot created\n")
  } else {
    cat("⚠ No significant fuel type pathways\n")
  }
  
  cat("\n=== DOT PLOT SUMMARY ===\n")
  cat("  Sex pathways:", nrow(sex_df_relevant_subset), "\n")
  cat("  Fuel pathways:", nrow(fueltype_df_relevant_subset), "\n")
}
```

---

# Session Information

```{r session-info}
cat("\n")
cat("═══════════════════════════════════════════════════════════\n")
cat("  RNA-SEQ ANALYSIS COMPLETE\n")
cat("═══════════════════════════════════════════════════════════\n\n")

cat("Output files saved to:\n")
cat("  Processed data:", here::here(PATH_DATA_PROCESSED), "\n")
cat("  Figures:", here::here(PATH_OUTPUT_FIGURES), "\n")
cat("  Tables:", here::here(PATH_OUTPUT_TABLES), "\n\n")

sessionInfo()
```